---
title: "Williams College Faculty and Graduates Data Manual"
author: "Karan Tibrewal and David Kane"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    fig_caption: true
---


```{r, echo = FALSE}
knitr::opts_chunk$set(tidy = FALSE,
                      echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
```


# Introduction

```{r}
library(williams)
library(dplyr)
data("graduates")
data("graduates_details")
```

The purpose of this document is to describe the procedure for the creating the faculty and graduate datasets. The data is collected from archives of the College's [course catalogs](https://catalog.williams.edu/archive/), from `r min(graduates$year)` to `r max(graduates$year)`. 

##Graduates

We begin with graduates. Our package provides two datasets: `graduates` and `graduates_details`. The `graduates` dataset provides the following variables: 

```{r}
names(graduates) %>% as.data.frame()
```

The `graduates_details` dataset expands on the above, and provides the following extra columns: 

```{r}
 names(graduates_details)[! names(graduates_details) %in%  names(graduates)] %>% as.data.frame()
```

For details on these variables, try `?graduates` and `?graduates_details`.    

The datasets are created by scraping data from the course catalogs which are saved as `.txt` files in the `inst/extdata` directory. 
These files follow the naming convention of `graduates-<YEAR (YYYY)>-<YEAR + 1 (YYYY)>`, where `YEAR` is the year 
for the relevant catalog. For example, the text file with information about graduates from 2000 is saved 
as `graduates-2000-2001.txt`. The reason for this awkward naming convention is that the graduates from (spring) 2000 are 
listed in the course catalog which comes out in fall of 2000, which is the catalog for the 2000-2001 academic year. 
For convenience, we only save the pages containing information about graduates from the course catalogs to these files. 

Let us look inside these text files. The following is an section from `graduates-2000-2001.txt`. 
We will use it as an example to demonstrate the data munging procedure for gradutes: 

```
Bachelor of Arts, Summa Cum Laude
*DoHyun Tony Chung, with honors in Political Economy
+*Rebecca Tamar Cover, with highest honors in Astrophysics
*Amanda Bouvier Edmonds
*Douglas Bertrand Marshall, with highest honors in Philosophy
+*Michelle Pacholec, with honors in Chemistry
*Grace Martha Pritchard, with honors in English
*Michael Vernon Ramberg
+*Taylor Frances Schildgen, with highest honors in Geosciences
*Qiang Sun
*Laura Susan Trice
*Max McMeekin Weinstein, with highest honors in Philosophy
*Catherine Anne Williams, with highest honors in History
Bachelor of Arts, Magna Cum Laude
David Scott Adams
*Julianne Humphrey Anderson
*Michael Zubrow Barkin
Robert Charles Blackstone
*Marlin Chu
*Sarah Ann Cohen, with highest honors in English
Mark Douglas Conrad
*Ellen Griswold Cook
*Mary Bowman Cummins
*Yana Dadiomova
```
### gather_graduates()

This function brings together the information on all Williams graduates for the years that we have downloaded the data in the package. It
also adds Latin honors, Phi Betta Kappa, Sigma Xi, and birth year information.

The first step is to read in the information from the text files. This leads to the creation of a data frame with a column called `raw.text` with a row for each line in the text files. Once the `raw.text` for graduates is available, we can immediately infer some data points:  

* **Latin Honors:** The information for graduates in the text files are grouped by their Latin honors. Notice the lines reading "Bachelor of Arts, Summa Cum Laude" and "Bachelor of Arts, Magna Cum Laude" in the above example. All graduates between these lines graduated with Summa Cum Laude. Similarly, all graduates between the lines "Bachelor of Arts, Magna Cum Laude" and "Bachelor of Arts, Cum Laude" graduated with Magna Cum Laude. As of such, we classify Latin honors for each graduate as one of Summa Cum Laude, Magna Cum Laude, Cum Laude, or NA.   
* **Phi Beta Kappa:** The course catalogs indicate membership of Phi Beta Kappa by a "*" before the graduate's name. 
* **Sigma Xi:** The course catalogs indicate membership of Sigma Xi by a "+" before the graduate's name. 
* **Birth Year:** We calculate the birth year using heuristic of undegraduate graduation year minus 22. 



```{r, echo = TRUE}
x <- gather_graduates()
x %>% as.data.frame() %>% head(15)
```


### add_graduate_names()

We infer `first.name`, `middle.name`, and `last.name` for graduates from the `raw.text` column from above. The names are always provided before the first comma. 

A big problem is that, although the vast majority of people have 3 names, hundreds have just 2 or 4 and a handful have 5, e.g., Alfonso Rodrigo González del Riego! One (so far!) even has 6: Chloë Iambe Naomi Illyria Feldman Emison. It is tricky dealing with last names that (obviously?) include more than one word, like González del Riego.

The first name is easy: it is the first word in the name. Ignoring the complication that last names might be multiple words, we (in some cases, incorrectly!) assume it to be the last word in the name. If there are other words between the first and last name, they are classified as the middle name. 

```{r, echo = TRUE}
x <- add_graduate_names(x)
x %>% dplyr::select(raw.text, first.name, middle.name, last.name) %>% as.data.frame() %>% head(15)
```

### add_graduate_honors

This function takes as input a data frame which includes the raw text associated with each graduate. It returns that data frame along with
six new columns (`honor.1`, `honor.2`, `honor`, `major.1`, `major.2`, and `major`) associated with department honors. Department honors are conferred on graduates who write a thesis during their senior year. 

Information about department honors is always provided after the first comma in the raw text. A graduate may write two theses, in which case, information about the two are seperated by a comma. 

* `honor.1` and `major.1`: These fields describe the first department honor conferred on the graduate as per the course catalogs. The `honor.1` field provides the distinction for the honor as one of "honors", "highest honors" , or NA. The `major.1` field provides the department under which the thesis was written. It is NA if the graduate has not written a thesis.     
* `honor.2` and `major.2`: These fields describe the second department honor conferred on the graduate as per the course catalogs. The `honor.2` field provides the distinction for the honor as one of "honors", "highest honors" , or NA. The `major.2` field provides the department under which the second thesis was written. It is NA if the graduate has not written a second thesis.
* `honor` and `major`: We simplify department information as the highest honor received by the graduate, and the corresponding major.   

```{r, echo = TRUE}
x <- add_graduate_honors(x)
x %>% dplyr::select(raw.text, honor.1, major.1, honor.2, major.2, honor, major) %>% as.data.frame() %>% head(15)
```

### add_gender_graduates

The course catalogs do not provide the gender of the graduates, so we estimate it by using the [gender](https://cran.r-project.org/web/packages/gender/gender.pdf) and [genderizeR](https://cran.r-project.org/web/packages/genderizeR/index.html) package. 

These packages predict gender on the basis of names. We first use the `gender` package to predict gender according to the first and middle names of graduates. Alas, the `gender` package focuses heavily on American names, and therefore, this approach is ineffective for non-American names like DoHyun and Qiang. 


To account for this shortcoming, we supplement the `gender` package with the `genderizeR` package which provides more global coverage. However, this is a paid service that limits free access. So, we used it once and then saved the results.[^1] 

[^1]: The results from genderizeR are saved as data frame in sys.rda in the R directory as genderizeR.graduates. While adding information about subsequent years, this data set must be regenerated.

```{r, echo = TRUE}
x <- add_gender_graduates(x)
x %>% dplyr::select(first.name, middle.name, gender) %>% as.data.frame() %>% head(15)
```

### add_race


We use the [wru](https://cran.r-project.org/web/packages/wru/wru.pdf) package to estimate race on the basis of last name. The package utilizes the Bayes' Rule to compute the posterior probability from Voter Registration Records of each racial category for any given individual's name. We predict the individual's race as the racial category with the highest prosterior probability. 

```{r, echo = TRUE}
x <- add_race(x)
x %>% dplyr::select(last.name, p_whi, p_bla, p_his, p_asi,p_oth,race) %>% as.data.frame() %>% head(15)
```

### Wraping Up

The functions described above are called from the `create_graduates` function, that takes in a logical parameter `complete`. Calling `create_graduates` with `complete = TRUE` produces the `graduates_details` data set. If `complete = FALSE`, `create_graduates` trims a dozen or so extraneous columns from `graduates_details` to produce the `graduates` dataset.

`graduates <- create_graduates(complete = FALSE)`             
`graduates_details <- create_graduates(complete = TRUE)`

### Adding More Data

In order to use the package to create the `graduates` and `graduates_details` datasets for subsequent years, follow the following steps: 

1. **Add information about graduates from the course catalogs to text files**. Follow the naming convention of `graduates-<YEAR (YYYY)>-<YEAR + 1 (YYYY)>`, where `YEAR` is the year for the relevant catalog. For example, the text file with information about graduates from 2000 is saved as `graduates-2000-2001.txt`. These files need to be saved in the `inst/extdata` directory.        

2. **Edit the text files to follow the munging syntax.** Such editing may often require getting your hands dirty with the text files. We describe the syntax below. 

```
Bachelor of Arts, Summa Cum Laude
<Information about graduates with Summa Cum Laude latin honors> 
Bachelor of Arts, Magna Cum Laude
<Information about graduates with Magna Cum Laude latin honors> 
Bachelor of Arts, Cum Laude
<Information about graduates with Cum Laude latin honors> 
Bachelor of Arts
<Information about graduates with no latin honors> 
```
Each line in these text files (except those, like "Bachelor of Arts, Summa Cum Laude", that serve as demarcations between different latin honors), contains information about a single graduate. These lines are organized as: 

```
<name>, <honors information>
```

Here, `<name>` is just the name of the graduate, for example, "DoHyun Tony Chung".     

`<honors information>` provides information about department honors earned by the graduate (if any). They are of the form `with <honors level> in <department>`. For example, "with honors in Political Economy" or "with highest honors in Mathematics". If the graduate has received honors from more than one department, the information about each is delimitted by a comma (`<information about department honor 1>, <information about department honor 2>`). For example, "with honors in Political Economy, with highest honors in Mathematics".      

Finally, add information about Phi Beta Kappa & Sigma Xi. Membership of Phi Beta Kappa is indicated by a "*" before the graduate's name, and that of Sigma Xi Kappa by a "+" before the graduate's name. 

Then, for example, information about a graduate takes the form: "*DoHyun Tony Chung, with honors in Political Economy". 

3. **Recreate the `genderizeR` dataset for new graduates, and save it in `sysdata.rda`**. For more information on how to achieve this, refer to the [genderizeR documentation](https://cran.r-project.org/web/packages/genderizeR/genderizeR.pdf).       

4. **Create and save the datasets**.  

`graduates <- create_graduates(complete = FALSE)`             
`graduates_details <- create_graduates(complete = TRUE)`
`save(graduates, file = .../data/graduates.RData)`
`save(graduates_details, file = .../data/graduates_details.RData)`


 


## Faculty

Next up are faculty. Our package provides two datasets: `faculty` and `faculty_details`. Let us begin by examining the `faculty_details` dataset. It has the following columns: 

```{r}
names(faculty_details) %>% as.data.frame()
```

For details on these variables, try `?faculty_details` on your R console.    

The datasets are created by scraping data from the Course Catalogs which are saved as `.txt` files in the `inst/extdata` directory. These files follow the naming convention of `faculty-<YEAR (YYYY)>-<YEAR + 1 (YYYY)>`, where `YEAR` is the year for the relevant catalog. For example, the text file with information about graduates from 2000 is saved as `faculty-2000-2001.txt`. For convenience, we only save the pages containing information about faculty from the course catalogs to these files. 

Let us look inside these text files. The following is an abstract from `faculty-2000-2001.txt`. We will use it as an example to demonstrate the data munging procedure for faculty: 

```
* Daniel P. Aalberts# Assistant Professor of Physics# B.S. (1989) M.I.T.; Ph.D. (1994) M.I.T.       
* Colin C. Adams# Francis Christopher Oakley Third Century Professor of Mathematics# B.S. (1978) M.I.T.; Ph.D. (1983) University of Wisconsin      
Elizabeth M. Adler# Assistant Professor of Biology# B.S. (1975) Antioch College; Ph.D. (1986) University of Pennsylvania      
* Laylah Ali# Assistant Professor of Art# B.A. (1991) Williams; M.F.A. (1994) Washington University     
Marsha I. Altschuler# Associate Professor of Biology# B.S. (1972) University of Rochester; Ph.D. (1979) Indiana University     
Rachael Aráuz# Visiting Assistant Professor of Art# B.A. (1991) Wellesley; M.A. (1995) University of Pennsylvania      
Henry W. Art# Samuel Fessenden Clarke Professor of Biology and Director of the Center for Environmental Studies# A.B. (1966) Dartmouth; Ph.D. (1971) Yale       
Jennifer Austin# Assistant Professor of Romance Languages# B.A. (1991) Earlham College; M.A. (1996) Cornell University       
Michael Awkward# Sterling A. Brown ’22 Visiting Professor of English#         
Jean L. Bacon# Assistant Professor of Sociology# B.A. (1983) Wesleyan; Ph.D. (1993) University of Chicago        
Duane A. Bailey# Associate Professor of Computer Science# B.A. (1982) Amherst College; Ph.D. (1988) University of Massachusetts        
Jon Bakija# Assistant Professor of Economics# B.A. (1990) Wesleyan; Ph.D. (1999) University of Michigan        
Lois Banta# Visiting Associate Professor of Biology# B.A. (1983) Johns Hopkins; Ph.D. (1990) California Institute of Technology       
David E. Barnard# Assistant Professor of Physical Education# B.A. (1981) Wesleyan; M.A.L.S. (1989) Wesleyan         
```
### gather_faculty()

This function brings together the information on all Williams faculty for the years that we have downloaded the data in the package. It
also adds information about faculty on leave.

The first step is to read in the information from the text files. This leads to the creation of a data frame with a column called `raw.text` with a row for each line in the text files. Once the `raw.text` column for faculty is available, we can immediately infer the leave status for each faculty. The course catalogs follow the following convention: 

* **Academic Year:** A single asteriks ("*") before the faculty's name.
* **First Semester:** Two asteriks ("**") before the faculty's name.
* **Second Semester:** Three asteriks ("***") before the faculty's name.


```{r, echo = TRUE}
x <- gather_faculty()
x %>% as.data.frame() %>% head(15)
```


### add_faculty_names()

This function takes as input a data frame which includes the raw text associated with each faculty member. It returns that data frame along with two new columns: first.name and last.name. 

For faculty, information about the name is always present before the first "#" in the `raw.text` column.  We will assume assume that the first word in the name is always the first name. Middle names (if present) seem to be collapsed into a single alphabet after the first name. All preceding words after the first (except collapsed middle names) in the name will be the last name.

Then, for example, for "Danial P. Alberts", we infer "Daniel" as first name, and "Alberts" as last name. Similarly, for "Jennifer Austin", "Jenifer" as the first name, and "Austin" as the last. Alas, for "Ju-Yu Scarlett Jang", we infer first name as "Ju-Yu", and last name as "Scarlett Jang", although "Scarlett" is probably a middle name. However, this complication seem to be exceptions, rather than the rule. 


```{r, echo = TRUE}
x <- add_faculty_names(x)
x %>% dplyr::select(raw.text, first.name, last.name) %>% as.data.frame() %>% head(15)
```

### add_faculty_degrees

This function takes as input a data frame which includes the `raw.text` associated with each faculty member. It returns that data frame along with six new columns: first.degree, first.degree.year, first.degree.school, last.degree, last.degree.year and last.degree.school.

The relevant information is provided after the last "#" in the `raw.text` column. There is a delimiting ";" between information about different degrees in the case where faculty have more than one degree.

```{r}
head(x$raw.text, 5)
```

We record the lowest degree earned by a faculty as `first.degree` , the conferring school as `first.degree.school`, and the corresponding year as `first.degree.year`. `last.degree`, `last.degree.year` and `last.degree.school` give the same context as above, but for the highest degree earned by the faculty. Note, in the case where the faculty has only one degree, the lowest and highest degree refer to the same. 


```{r, echo = TRUE}
x <- add_faculty_degrees(x)
x %>% dplyr::select(first.degree, first.degree.year, first.degree.school, last.degree, last.degree.year, last.degree.school) %>% as.data.frame() %>% head(15)
```

### add_faculty_titles

This function takes as input a data frame which includes the raw text associated with each faculty member. It returns that data frame along with 2 new columns: for the facutly's title and faculty's rank.

The facultys' job title is always present between the first and second '#'s in the `raw.text` column. It easy to grab using `stringr::str_split`. Once we have the title in place, we can infer the rank: 

* **Assistant Professor** If the words "Assistant Professor" appears on the job title.
* **Assosciate Professor** If the words "Assosciate Professor" appears on the job title.
* **Professor** If the word "Professor" appears on the job title, but none of the above.
* **Fellow:** If the word "Fellow" appears on the job title, but none of the above.
* **Librarian:** If the word "Librarian" appears on the job title, but none of the above.
* **Instructor:** If the word "Instructor" appears on the job title, but none of the above.
* **Lecturer:** If the word "Lecturer" appears on the job title, but none of the above.
* **Artist-in-Residence:** If the word "Artist-in-Residence" appears on the job title, but not "Professor", but none of the above. 

```{r}
x <- add_faculty_titles(x)
x %>% as.data.frame() %>% dplyr::select(raw.text, title) %>% head(15)
```

###add_faculty_department

The course catalogs do not directly provide the department for faculty. Instead, we try to infer the department by a faculty's title. For some, this is easy. For example, "Daniel P. Aalberts, Assistant Professor of Physics" definitely belongs to Physics, and "Colin C. Adams, Francis Christopher Oakley Third Century Professor of Mathematics" to Mathematics. In order to accomplish this, we enlist the various Williams academic departments in text file `departments.txt` in `inst/extdata` directory, and use `stringr::str_detect` on the title. 

Others are more complicated. For example, consider "Bernadette Brooten, Croghan Bicentennial Visiting Professor in Biblical
and Early Christian Studies, Spring Semester". Although, from Bernadette Brooten's title, it is fairly obvious that she is a Professor of Religion, her department cannot be inferred from the naive procedure with which we infer Colin Adams' or Daniel Alberts' department. We need to handle these cases separately.    

Alas, others are, still, more complicated. "Jennifer Austin, Assistant Professor of Romance Languages" is a case in point: it is obvious that she teaches Romance Languages, but does she teach French, Italian, or Spanish? These cases need to be handled explicitly by hand.  

```{r}
x <- add_faculty_department(x)
x %>% as.data.frame() %>% dplyr::select(first.name, last.name, title, department) %>% head(15)
```


###add_gender_faculty

The course catalogs do not provide the gender of the faculty, so we estimate it by using the [gender](https://cran.r-project.org/web/packages/gender/gender.pdf) and [genderizeR](https://cran.r-project.org/web/packages/genderizeR/index.html) package. 

These packages predict gender on the basis of names. We first use the `gender` package to predict gender according to the first names of faculty. Alas, the `gender` package focuses heavily on American names, and therefore, this approach is ineffective for non-American names like Aamir and Amar. 

To account for this shortcoming, we supplement the `gender` package with the `genderizeR` package which provides more global coverage. However, this is a paid service that limits free access. So, we used it once and then saved the results. [^2] 

[^2]: [The results from genderizeR are saved as data frame in sys.rda in the R directory as genderizeR.faculty. While adding information about subsequent years, this data set must be regenerated.]


```{r, echo = TRUE}
x$birth.year <- x$first.degree.year - 22
x <- add_gender_faculty(x)
x %>% dplyr::select(first.name, gender) %>% as.data.frame() %>% head(15)
```


###add_race

We use the [wru](https://cran.r-project.org/web/packages/wru/wru.pdf) package to estimate race on the basis of last name. The package utilizes the Bayes' Rule to compute the posterior probability from Voter Registration Records of each racial category for any given individual's name. We predict the individual's race as the racial category with the highest prosterior probability. 

```{r, echo = TRUE}
x <- add_race(x)
x %>% dplyr::select(last.name, p_whi, p_bla, p_his, p_asi,p_oth,race) %>% as.data.frame() %>% head(15)
```


###Wrapping Up

The functions described above are called from the `create_faculty` function, that takes in a logical parameter `complete`. Calling `create_faculty` with `complete = FALSE` produces the `faculty_details` data set. If  `complete = TRUE` several intermediarry columns are appended to `faculty_details`. 

`faculty_details <- create_faculty(complete = FALSE)`


**Creating `faculty` dataset:**     

Now that we understand how to create the `faculty_details` from scratch, let us consider the `faculty` dataset. Simply put, `faculty` summarizes `faculty_details` for the current year. We can create it by using the `create_annual_faculty` function. 

`faculty <- create_annual_faculty(faculty_details)`

In addition to filtering the `faculty_detais` dataset for the current year, the `create_annual_faculty` also consolidates any information about faculty that is not present in the latest course catalog (and hence, also in the rows containing information about the latest year). For example, if the undergraduate degree for a faculty is not listed in the course catalog for 2016, but was listed in the 2015 course catalog, `create_annual_faculty` will transfer this information onto the lower columns.        


### Adding More Data

Steps to use the package to create the `faculty` and `fauclty_details` datasets for subsequent years: 

1. **Add information about faculty from the course catalogs to text files**. Follow the naming convention of `faculty-<YEAR (YYYY)>-<YEAR + 1 (YYYY)>`, where `YEAR` is the year for the relevant catalog. For example, the text file with information about faculty from 2000 is saved as `faculty-2000-2001.txt`. These files need to be saved in the `inst/extdata` directory. 

2. **Edit the text files to follow the munging syntax.** Such editing may often require getting your hands dirty with the text files. We describe the syntax below. 

Each line in these text files contains information about a single faculty. These lines are organized as: 

```
<leave information> <name> # <job title> # <degree information> 
```


Here, `<leave information` is described by asteriks ("*") before the faculty's name.  Leave for the Academic Year is denoted by a single asteriks ("*"), while that for the first semester, by two asteriks ("**"), and that for the second semester, by three asteriks ("***").

`<name>` is just the name of the faculty, for example, "Daniel P. Alberts".    

`<job title>` is the job title of the faculty, which is provided in the course catalogs. For example, "Professor of Mathematics", or "Sterling A. Brown ’22 Visiting Professor of English". 

`degree information` provides information about the faculty's academic career. They are of the form `<degree> (<year>) <conferring school>`, where `<degree>` is the name of the degree, `<year>` is the year in which the degree was conferred, and `<conferring school>`, as the name suggests, in the school that conferred the degree on the faculty. For example, `<degree information>` about a faculty, may take the form "B.S. (1978) M.I.T.; Ph.D. (1983) University of Wisconsin". Notice that if the faculty has earned more than one degree, than information about each is delimitted by a semi-colon.     

Then, for example, information about a faculty takes the form: "* Daniel P. Aalberts# Assistant Professor of Physics# B.S. (1989) M.I.T.; Ph.D. (1994) M.I.T." 

3. **Recreate the `genderizeR` dataset for new faculty, and save it in `sysdata.rda`**. For more information on how to achieve this, refer to the [genderizeR documentation](https://cran.r-project.org/web/packages/genderizeR/genderizeR.pdf).       

4. **Create and save the datasets.**  

`faculty_details <- create_faculty(complete = FALSE)`             
`faculty <- create_annual_faculty(faculty_details)`
`save(faculty, file = .../data/faculty.RData)`
`save(faculty_details, file = .../data/faculty_details.RData)`




