---
title: "Williams College Faculty and Graduates"
author: "Karan Tibrewal and David Kane"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
---


```{r, echo = FALSE}
knitr::opts_chunk$set(tidy = FALSE,
                      echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
```


## Introduction

This document provides an overview of the data available in the wiliamsmetrics package. We have data on the faculty and graduates of Williams College from the 1999--2000 academic year through 2014--2015.

```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(williamsmetrics)
data(graduates_details)
x <- graduates_details
```

## Faculty

We have not completed the collection of faculty data.

## Graduates

There were `r dim(x)[1]` graduates of Williams from `r min(x$year)` through `r max(x$year)`. The number of graduates averages about `r round(mean(table(x$year)))` per year. The largest class was `r as.numeric(tail(sort(table(x$year)), n = 1))` in 
`r as.numeric(names(tail(sort(table(x$year)), n = 1)))`. The smallest was 
`r as.numeric(head(sort(table(x$year)), n = 1))` in
`r as.numeric(names(head(sort(table(x$year)), n = 1)))`. 

### Names

Names are interesting. The graduate with the longest name is 
`r x %>% select(year, full.name) %>% filter(min_rank(-nchar(full.name)) < 2) %>% as.character() %>% tail(1)` 
from the class of 
`r x %>% select(year, full.name) %>% filter(min_rank(-nchar(full.name)) < 2) %>% as.character() %>% head(1)`. The students with the shortest names are:

```{r, echo = FALSE}
x %>% select(year, full.name) %>% filter(min_rank(nchar(full.name)) < 2) %>% as.data.frame()
```

The students with the longest last name and first name are:

```{r, echo=FALSE}
x %>% select(year, full.name, first.name, last.name) %>% filter(min_rank(-nchar(first.name)) < 2 | min_rank(-nchar(last.name)) < 2) %>% select(-first.name, -last.name) %>% as.data.frame()
```

The students with at least 5 separate parts to their full names are:

```{r, echo=FALSE}
x[str_split(x$full.name, " ", simplify = TRUE)[,5] != "",] %>% select(year, full.name) %>% as.data.frame()
```

### Gender

The course catalogs do not provide the gender of the graduates, so we estimate it by using the **gender** package. 
`r tufte::margin_note("At some point, we hope to explore a more advanced approach: Kamil Wais (2016). genderizeR: Gender Prediction Based on First Names. R package version 2.0.0. https://CRAN.R-project.org/package=genderizeR")` Our gender variable looks like:

```{r}
table(x$gender, exclude = NULL)
```

The large number of NAs is a result of the gender package's focus on American names. Consider the most common problem cases:

```{r}
x %>% filter(is.na(gender)) %>% select(first.name, gender) %>% group_by(first.name, gender) %>% summarize(count = n()) %>% arrange(desc(count)) %>% as.data.frame() %>% head(12)
```

Most of the NAs for gender are due to unique first names, names that appear only once amoung the `r dim(x)[1]` graduates. Eventually, we might fix some of these by hand, especially those --- like José --- for which we know the correct gender. There might also be simple ways to add international names. But, for now, we see no reason why these NA should bias any of the later analysis. 

```{r}
x %>% select(year, gender) %>% count(year, gender) %>% mutate(percentage = n/sum(n)) %>% ggplot(aes(year, percentage, color = gender)) + geom_line()
```

### Race

We use the wru package to estimate race. The totals for the entire data set are:

```{r}
table(x$race, exclude = NULL)
```

Judging by the the College's public statements and Common Data Set releases, we are significantly overestimating the percentage of white students.

```{r}
round(prop.table(table(x$race)), 3)
```

The "Other" category includes only a handful of students:

```{r}
x %>% filter(race == "Other") %>% select(full.name, race) %>% as.data.frame()
```

There is no publicly available source for the racial breakdown of the graduating class. (True? Check NEPS.) The 2015--2016 Common Data Set [reports](https://provost.williams.edu/files/williams_cds_1516_w_tuition.pdf) that, among all degree-seeking undergraduates at Williams, `r round(243/2065, 2)` Asian, `r round(157/2065, 2)` were 
Black, `r round(261/2065, 2)` Hispanic, `r round(1106/2065, 2)` White, and `r round((161 + 4 + 133)/2065, 2)` in other categories, including Nonresident Aliens, American Indian or Two-or-more races.

Our estimates for non-White students are too low. First, we don't even find the same number of Asian/Black/Hispanic students are we know to be at Williams among the US citizen portion of the population. Second, we categorize some/many of the students in the other categories (most importantly the international and multi-racial students) as being in non-White categories even though they are not so classified by Williams.

As an example, we classify Paige Lauren Whidbee '16 as Black. She is the only Phi Beta Kappa member of the class of 2016 that we so classify. However, according to this [photo](http://ephsports.williams.edu/sports/wgolf/2014-15/bios/whidbee_paige_2xmt?view=bio), she probably isn't African-American.





```{r}
x %>% select(year, race) %>% count(year, race) %>% mutate(percentage = n/sum(n)) %>% ggplot(aes(year, percentage, color = race)) + geom_line()
```
