---
title: "Williams College Faculty and Graduates"
author: "Karan Tibrewal and David Kane"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
---


```{r, echo = FALSE}
knitr::opts_chunk$set(tidy = FALSE,
                      echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
```


## Introduction

This document provides an overview of the data available in the wiliamsmetrics package. We have data on the faculty and graduates of Williams College from the 1999--2000 academic year through 2014--2015.

```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(williamsmetrics)
data(graduates_details)
x <- graduates_details
```


## Graduates

There were `r dim(x)[1]` graduates of Williams from `r min(x$year)` through `r max(x$year)`. The number of graduates averages about `r round(mean(table(x$year)))` per year. The largest class was `r as.numeric(tail(sort(table(x$year)), n = 1))` in 
`r as.numeric(names(tail(sort(table(x$year)), n = 1)))`. The smallest was 
`r as.numeric(head(sort(table(x$year)), n = 1))` in
`r as.numeric(names(head(sort(table(x$year)), n = 1)))`. 

### Names

Names are interesting. The graduate with the longest name is 
`r x %>% select(year, full.name) %>% filter(min_rank(-nchar(full.name)) < 2) %>% as.character() %>% tail(1)` 
from the class of 
`r x %>% select(year, full.name) %>% filter(min_rank(-nchar(full.name)) < 2) %>% as.character() %>% head(1)`. The students with the shortest names are:

```{r, echo = FALSE}
x %>% select(year, full.name) %>% filter(min_rank(nchar(full.name)) < 2) %>% as.data.frame()
```

The students with the longest last name and first name are:

```{r, echo=FALSE}
x %>% select(year, full.name, first.name, last.name) %>% filter(min_rank(-nchar(first.name)) < 2 | min_rank(-nchar(last.name)) < 2) %>% select(-first.name, -last.name) %>% as.data.frame()
```

The students with at least 5 separate parts to their full names are:

```{r, echo=FALSE}
x[str_split(x$full.name, " ", simplify = TRUE)[,5] != "",] %>% select(year, full.name) %>% as.data.frame()
```

### Gender

The course catalogs do not provide the gender of the graduates, so we estimate it by using the **gender** package. 
`r tufte::margin_note("At some point, we hope to explore a more advanced approach: Kamil Wais (2016). genderizeR: Gender Prediction Based on First Names. R package version 2.0.0. https://CRAN.R-project.org/package=genderizeR")` Our gender variable looks like:

```{r}
table(x$gender, exclude = NULL)
```

The large number of NAs is a result of the gender package's focus on American names. Consider the most common problem cases:

```{r}
x %>% filter(is.na(gender)) %>% select(first.name, gender) %>% group_by(first.name, gender) %>% summarize(count = n()) %>% arrange(desc(count)) %>% as.data.frame() %>% head(12)
```

Most of the NAs for gender are due to unique first names, names that appear only once amoung the `r dim(x)[1]` graduates. Eventually, we might fix some of these by hand, especially those --- like José --- for which we know the correct gender. There might also be simple ways to add international names. But, for now, we see no reason why these NA should bias any of the later analysis. 

```{r}
x %>% select(year, gender) %>% count(year, gender) %>% mutate(percentage = n/sum(n)) %>% ggplot(aes(year, percentage, color = gender)) + geom_line()
```

### Race

We use the wru package to estimate race. The totals for the entire data set are:

```{r}
table(x$race, exclude = NULL)
```

Judging by the the College's public statements and Common Data Set releases, we are significantly overestimating the percentage of white students.

```{r}
round(prop.table(table(x$race)), 3)
```

The "Other" category includes only a handful of students:

```{r}
x %>% filter(race == "Other") %>% select(full.name, race) %>% as.data.frame()
```

There is no publicly available source for the racial breakdown of the graduating class. (True? Check NEPS.) The 2015--2016 Common Data Set [reports](https://provost.williams.edu/files/williams_cds_1516_w_tuition.pdf) that, among all degree-seeking undergraduates at Williams, `r round(243/2065, 2)` Asian, `r round(157/2065, 2)` were 
Black, `r round(261/2065, 2)` Hispanic, `r round(1106/2065, 2)` White, and `r round((161 + 4 + 133)/2065, 2)` in other categories, including Nonresident Aliens, American Indian or Two-or-more races.

Our estimates for non-White students are too low. First, we don't even find the same number of Asian/Black/Hispanic students are we know to be at Williams among the US citizen portion of the population. Second, we categorize some/many of the students in the other categories (most importantly the international and multi-racial students) as being in non-White categories even though they are not so classified by Williams.

As an example, we classify Paige Lauren Whidbee '16 as Black. She is the only Phi Beta Kappa member of the class of 2016 that we so classify. However, according to this [photo](http://ephsports.williams.edu/sports/wgolf/2014-15/bios/whidbee_paige_2xmt?view=bio), she probably isn't African-American.





```{r}
x %>% select(year, race) %>% count(year, race) %>% mutate(percentage = n/sum(n)) %>% ggplot(aes(year, percentage, color = race)) + geom_line()
```


## Faculty

```{r}
data("faculty")
data("faculty_details")
```

There were `r dim(faculty_details)[1]` faculty at Williams from `r min(faculty_details$year)` through `r max(faculty_details$year)`. The number of faculty averages about `r round(mean(table(faculty_details$year)))` per year. The largest class was `r as.numeric(tail(sort(table(faculty_details$year)), n = 1))` in 
`r as.numeric(names(tail(sort(table(faculty_details$year)), n = 1)))`. The smallest was 
`r as.numeric(head(sort(table(faculty_details$year)), n = 1))` in
`r as.numeric(names(head(sort(table(faculty_details$year)), n = 1)))`. 

###Gender

We infer the gender for faculty using the `gender` and `genderizeR` package, as in the case of graduates. The gender variable for faculty currently employed at Williams looks like the following: 

```{r}
round(prop.table(table(faculty$gender)), 3)
```

These figures are similar to those [reported](https://provost.williams.edu/files/williams_cds_1617.pdf) by the Williams 2016-2017 Common Data Set. 



```{r}
faculty_details %>% select(year, gender) %>% count(year, gender) %>% mutate(percentage = n/sum(n)) %>% ggplot(aes(year, percentage, color = gender)) + geom_line() + labs(title = "Comosition of Williams Faculty by Gender", x = "Year", y = "Percentage")
```

###Race

As with graduates, we use the wru package to estimate race. The totals for the year `r paste(faculty$year[1], faculty$year[1] + 1, sep = ',')`:

```{r}
tab <- table(faculty$race, exclude = NULL)
```

Once again, we seem to be overestimating the number of white faculty. Information about racial breakdown of faculty at Williams is unavailable, but the Williams 2016-2017 Common Data Set [reports]((https://provost.williams.edu/files/williams_cds_1617.pdf)) that the percentage of minority faculty is `r round(60/278, digits = 2)`, while we predict it to be `r round(sum(tab[1:4])/ sum(tab), digits = 2)`. The individual breakdown is: 

```{r}
round(prop.table(table(faculty$race)), 3)
```

For the most part, our prediction for Asian and Hispanic faculty seems to be correct. In fact, in 2016-2017, James A. Blair is the only faculty member misclasified as "Asian", and Janis Bravo and Peter Monteil were the two misclassified as Hispanic. 

The problem lies in differentiating between White and Black faculty. In 2016-2017, about half of faculty classified as "Black", are not in fact, African-American. For example, consider [Richard De Veaux](https://media.licdn.com/mpr/mpr/shrinknp_200_200/p/8/000/253/054/273cd6d.jpg). 



```{r}
faculty_details %>% select(year, race) %>% count(year, race) %>% mutate(percentage = n/sum(n)) %>% ggplot(aes(year, percentage, color = race)) + geom_line() + labs(title = "Comosition of Williams Faculty by Race", x = "Year", y = "Percentage")
```

###Undegraduate Degree

The course catalogs provide information about academic degrees earned by faculty. For a given faculty, we infer their undergraduate, or lowest degree, as the first degree we have information about. 

The `faculty` and `faculty_details` datasets provide information about faculty's undergraduate college degree (as defined above), the conferring school, and the year the degree was received. 

The top 10 most common undergraduate programs that faculty attended are: 

```{r}
college.summary <- dplyr::filter(faculty, !is.na(first.degree.school)) %>% dplyr::group_by(first.degree.school) %>% summarise(n = n())
college.summary <- college.summary[with(college.summary, order(-n)), ]
college.summary[1:10, ] %>% as.data.frame()
```


###Highest Degree

The `faculty` and `faculty_details` datasets provide information about the highest degree earned by faculty (as inferred by course catalogs), the conferring school, and the year the degree was received. 

The top 10 most common graduate programs that faculty attended are: 

```{r}
college.summary <- dplyr::filter(faculty, !is.na(last.degree.school)) %>% dplyr::group_by(last.degree.school) %>% summarise(n = n())
college.summary <- college.summary[with(college.summary, order(-n)), ]
college.summary[1:10, ] %>% as.data.frame()
```

##Title

The college's course catalogs provide the job title for faculty members. They are interesting because, as we will see later, they give us information about facultys' rank and department. 

The five most common titles are: 

```{r}
 sort(table(faculty$title), decreasing = T)[1:5] %>% as.data.frame() %>% dplyr::select(title = Var1, count = Freq)
```


##Rank

```{r}
x <- select(faculty_details, start.year = year, first.name, last.name, start.rank = rank)
faculty_start.rank <- dplyr::left_join(faculty, 
                            x, 
                            by = c("first.name", "last.name", "start.year"))
```

We infer the rank of faculty as one of Professor, Assistant Professor, Lecturer, Associate Professor, Artist-in-Residence, or Fellow from their job title. For example, we classify the rank of "Daniel P. Aalberts, Assistant Professor of Physics" as "Assistant Professor", and  that for "Irma Mayorga, Gauius Charles Bolin Fellow" as "Fellow". 

Among faculty currently employed at Williams, the rank variable looks like: 

```{r}
table(faculty_start.rank$rank)
```
 
```{r}
grouped <- dplyr::filter(faculty, !is.na(rank)) %>% dplyr::group_by(race, rank)
summarised <- dplyr::summarise(grouped, n = n())
ggplot(data = summarised, aes(x = rank, y = n, fill = race)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Faculty Rank by Race", x = "Rank", y = "Count") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
grouped <- dplyr::filter(faculty, !is.na(rank)) %>% dplyr::group_by(gender, rank)
summarised <- dplyr::summarise(grouped, n = n())
ggplot(data = summarised, aes(x = rank, y = n, fill = gender)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Faculty Rank by Gender", x = "Rank", y = "Count") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
assistant_prof <- faculty_start.rank[which(faculty_start.rank$rank == "Professor" & 
                                           faculty_start.rank$start.rank == "Assistant Professor"), ] 
assoc_prof <- faculty_start.rank[which(faculty_start.rank$rank == "Professor" & 
                                           faculty_start.rank$start.rank == "Assistant Professor"), ] 
prof <- faculty_details[which(faculty_details$rank == "Professor"), ] %>% 
       dplyr::filter(!duplicated(paste(first.name, last.name))) %>%
       dplyr::select(first.name, last.name, promotion.year = year)
assistant_prof <- dplyr::left_join(assistant_prof, prof, by = c("first.name", "last.name"))
assoc_prof <- dplyr::left_join(assoc_prof, prof, by = c("first.name", "last.name"))

assistant_prof$promotion_time <- assistant_prof$promotion.year - assistant_prof$start.year
assoc_prof$promotion_time <- assoc_prof$promotion.year - assoc_prof$start.year
```


We find that of the `r total = sum(faculty_start.rank$rank == "Professor", na.rm = T)` `r total` faculty members with rank `Professor` currently employed at Williams, `r round(sum(faculty_start.rank$rank == "Professor" & faculty_start.rank$start.rank == "Assistant Professor", na.rm = T)/total * 100, 2)`% were first employed as `Assistant Professor` and at average, took `r round(mean(assistant_prof$promotion_time), 2)` years to be promoted. `r round(sum(faculty_start.rank$rank == "Professor" & faculty_start.rank$start.rank == "Associate Professor", na.rm = T)/total * 100, 2)`% were first employed as `Associate Professor`, and at average, took `r round(mean(assoc_prof$promotion_time), 2)` years to be promoted.


###Department

```{r}
faculty.na <- faculty_details[which(!is.na(faculty_details$department)), ]
grouped <-  group_by(faculty.na, year, department)
summary.df <- summarise(grouped, number_of_faculty = n())
summary.df.top10 <- summary.df[with(summary.df, order(-number_of_faculty)), ] %>% filter(year == max(summary.df$year))
summary.df.top10 <- summary.df.top10[1:10, ]
```



The course catalogs do not directly provide the department for faculty. Instead, we try to infer the department by a faculty's title. For some, this is easy. For example, "Daniel P. Aalberts, Assistant Professor of Physics" definitely belongs to Physics, and "Colin C. Adams, Francis Christopher Oakley Third Century Professor of Mathematics" to Mathematics.    

Others are more complicated. For example, consider "Bernadette Brooten, Croghan Bicentennial Visiting Professor in Biblical
and Early Christian Studies, Spring Semester". Although, from Bernadette Brooten's title, it is fairly obvious that she is a Professor of Religion, her department cannot be inferred from the naive procedure with which we infer Colin Adams' or Daniel Alberts' department. We need to handle these cases separately.    

Alas, others are, still, more complicated. "Jennifer Austin, Assistant Professor of Romance Languages" is a case in point: it is obvious that she teaches Romance Languages, but does she teach French, Italian, or Spanish? These cases need to be handled explicitly.       


According to our findings, the top 10 departments in terms of faculty employed for the latest year we have data for (`r paste(max(faculty_details$year), max(faculty_details$year) + 1, sep = '-')`) are: 

```{r}
summary.df.top10  %>% as.data.frame()
```

```{r, fig.cap="Top 10 departments by number of faculty employed"}
filter.summary <- summary.df %>% filter(department %in% summary.df.top10$department)
p <- ggplot(data = filter.summary, aes(x = year, y = number_of_faculty, color = department)) + geom_line() + labs(x = "Year", y = "Number of faculty", title = "Number of Faculty by Department")

p 
```

###Birth Year

The course catalogs provide the year a faculty member graduated from college. Assuming (perhaps incorrectly in some cases) that she is 22 when she graduates, we infer her birth year.   

```{r}
faculty$age <- 2017 - faculty$birth.year
oldest <- paste(faculty$first.name[which(faculty$age == max(faculty$age, na.rm = T))], faculty$last.name[which(faculty$age == max(faculty$age, na.rm = T))], sep = ' ')
youngest <- paste(faculty$first.name[which(faculty$age == min(faculty$age, na.rm = T))], faculty$last.name[which(faculty$age == min(faculty$age, na.rm = T))], sep = ' ')
```

We find that the oldest faculty currently in the college is `r oldest[1]`. The youngest faculty currently in the college is `r youngest[1]`. 
The age variable for faculty currently employed at Williams looks like:

```{r}
summary(faculty$age)
```

```{r}
faculty$age <- 2017 - faculty$birth.year
qplot(faculty$age, geom="histogram", fill = I("purple"), color = I("black") ) + labs(title = "Distribution of Age of Faculty currently employed at Williams", y = "Count", x = "Age")

```


The average age for both male and female faculty has constantly decreased over the years (with the only exception of slight increases during 2007 to 2009. This is, perhaps, expected in view of the financial crisis when Williams had to considerably downsize its faculty. Presumably, newer, and hence, younger faculty, were disproportionately axed).   

`r tufte::margin_note("Note: College catalogs for 2015-2016 do not contain information about faculty's age." )`

```{r, fig.cap="Average age of faculty by gender"}
faculty_details$age <- 2017 - faculty_details$birth.year
grouped.year <- dplyr::group_by(faculty_details, year, gender)
summary.year <- dplyr::summarise(grouped.year, avg_age = mean(age, na.rm = T))
ggplot(data = summary.year, aes(x = year, y = avg_age, fill = gender)) + geom_bar(stat = "identity", position = "dodge") + labs(title = "Average Age of Faculty by Gender", x = "Year", y = "Average Age")
```

The average age of all but Black faculty has fallen steadily over the years. The average age of Black faculty increased during the first few years of 2000, and then steadily decreased to reach similar levels as other races. 

`r tufte::margin_note("Note: College catalogs for 2015-2016 do not contain information about faculty's age." )`

```{r}
grouped.year <- dplyr::group_by(faculty_details, year, race)
summary.year <- dplyr::summarise(grouped.year, avg_age = mean(age, na.rm = T))
ggplot(data = summary.year, aes(x = year, y = avg_age, fill = race)) + geom_bar(stat = "identity", position = "dodge") + labs(title = "Average Age of Faculty by Race", x = "Year", y = "Average Age")
```

###Start Year

We infer a faculty's `start.year`, or the first year they joined Williams, as the year of the first catalog that lists them as faculty of the college. 

`r tufte::margin_note(paste("Note: this inference is correct for all faculty, but for whom we infer the year to be the first year we have data about, that is ", min(faculty_details$year, na.rm = T), ". This is because these faculty may have very well started their career at Williams earlier than ", min(faculty_details$year, na.rm = T), "and we just don't have any data to know precisely. For these faculty, the start.year should be perhaps more accurately interpreted as <= ", min(faculty_details$year, na.rm = T), ".", sep = " "))`

The `start.year` variable gives us insight into how long the current faculty at Williams have been around: 
```{r}
faculty$term <- 2017 - faculty$start.year
summary(faculty$term)
```
```{r}
qplot(faculty$term, geom="histogram", fill = I("purple"), color = I("black") ) + labs(title = "Length of Tenure of Williams College faculty", y = "Count", x = "Years") + scale_y_continuous(limits = c(0, 100))
```


```{r}
dplyr::filter(faculty, !is.na(gender)) %>% ggplot(aes(term, fill = gender)) + geom_histogram(aes(y = ..density..), position = 'dodge') + labs(title = "Length of Tenure of Williams College Faculty by Gender", y = "Count", x = "Years")

```


```{r}
dplyr::filter(faculty, !is.na(gender)) %>% ggplot(aes(term, fill = race)) + geom_histogram(aes(y = ..density..), position = 'dodge') + labs(title = "Length of Tenure of Williams College Faculty by Race", y = "Count", x = "Years")

```
